Emulsions Stabilized by Magnetic Ellipsoidal Particles: A Lattice Boltzmann Study

### 2025-04-14 chapters/paper3.tex
Removed some repetition in chapter 7

While the rheology of emulsions stabilized by spherical particles has been extensively characterized, ellipsoidal particles introduce additional complexity through interfacial locking and anisotropic drag, 
which can suppress coalescence and resist flow differently under shear \cite{madivala_self-assembly_2009, daware_emulsions_2015, sun_assembly_2013}. Under shear, ellipsoidal particles 
can promote shear banding, wherein particle-fluid interactions create localized variations in velocity profiles \cite{xu_relation_2013}. Furthermore, suspensions of magnetically responsive particles, 
such as ferrofluids, exhibit a pronounced increase in viscosity due to field-induced self-assembly that restricts fluid flow \cite{qiao_magnetorheological_2012}. 
Theoretical predictions and experimental observations have identified that orientational ordering of anisotropic particles in the direction of shear reduces viscosity by minimizing hydrodynamic drag and interfacial 
disruption, allowing the fluid to flow more easily along the direction of least resistance.\cite{vermant_flow-induced_2005, xu_relation_2013}.

@@

To probe the rheological behavior of bijels stabilized by ellipsoidal particles, we analyze their response under constant shear with particular focus on the influence of interfacial particle ordering 
and magnetic field application. The initial bijel structures are obtained from previous simulations, consisting of bicontinuous morphologies stabilized by either prolate or oblate magnetic 
ellipsoids at a particle volume fraction of $\phi_p = 0.15$, generated with and without an applied magnetic field of dimensionless strength $\bar{B} = 1$. To evaluate the combined and decoupled effects 
of field history and applied shear, we systematically vary the magnetic field configuration, yielding four distinct processing pathways: 
$\bar{B}:0 \rightarrow 0$, $\bar{B}:0 \rightarrow 1$, $\bar{B}:1 \rightarrow 0$, and $\bar{B}:1 \rightarrow 1$. These setups enable us to isolate the contributions of pre-aligned versus dynamically aligning 
particles during flow. Shear is imposed via a range of dimensionless shear capillary numbers, defined as $Ca_s = \frac{\dot{\gamma} R_s \eta_f}{\sigma}$, where $\dot{\gamma}$ is the applied shear rate, 
$R_s = 7.9$ is the equivalent spherical radius of the ellipsoids, $\eta_f$ is the dynamic viscosity of the continuous phase, and $\sigma$ is the interfacial tension. 
\cite{frijters_effects_2012, yang_capillary_2022} We explore five shear conditions spanning $Ca_s = 1 \times 10^{-4}$ to $1 \times 10^{-6}$, capturing the transition from strong deformation to 
near-equilibrium interfacial response.

@@

Particle ordering is one factor that affects the rheological behavior of suspensions, as aligned particles reduce particle-fluid interactions, reducing the
induced shear stress. In bijels, where particles are adsorbed at the fluid-fluid interface, interfacial confinement may significantly 
influence the degree of slip and the onset of yielding. Prior studies have shown that bijels exhibit shear-thinning behavior consistent with the Herschel-Bulkley fluid model, 
indicating a non-Newtonian response governed by both interfacial structure and particle dynamics \cite{macmillan_rheological_2019}. To investigate these effects in 
the presence of magnetic fields and varying degrees of particle ordering, we apply constant shear to the bijel templates and plot the resulting time evolution of 
the shear stress, as shown in Figure~\ref{fig:stress_time}.

@@

Figure \ref{fig:stress_time} demonstrates that a definitive steady-state shear stress is not reached within the simulated timeframes across all conditions. 
This absence of steady-state behavior is attributed to the continued coarsening of the bicontinuous domains and the progressive reorientation of interfacial 
particles under shear \cite{tavacoli_novel_2011, macmillan_rheological_2019}. Tavacoli et al. showed that mechanical perturbation of bijels, such as the 
insertion of a needle to measure yield stress, resulted in irreversible structural deformation, particularly in ethanediol/nitromethane systems 
\cite{tavacoli_novel_2011}. Similarly, Macmillan et al. observed that repeated shearing of bijels synthesized by direct mixing led to degraded mechanical 
performance over time, suggesting a link between structural evolution and rheological response \cite{macmillan_rheological_2019}. To characterize the rheology 
of these evolving systems, we identify pseudo-steady-state regions within the stress-time profiles, defined as intervals where the rate of change of shear 
stress is minimal and persists for the longest continuous duration. Within these regions, the stress response is fitted to the applied shear rate using a 
Herschel-Bulkley model:

@@

In the Herschel-Bulkley model, $\sigma$ denotes the computed shear stress, $\sigma_y$ is the yield stress, $K$ represents the flow consistency index, 
and $n$ is the flow index characterizing the degree of shear thinning. In certain cases, simulations conducted at the lowest shear capillary number 
($Ca_s = 10^{-6}$) exhibited insufficient stress resolution or high variability, and were excluded from model fitting where appropriate. For each case, 
$\sigma$ is extracted from the pseudo-steady-state region of the stress-time curves, while the corresponding shear rate $\dot{\gamma}$ is derived from 
the imposed capillary number, $Ca_s$. The resulting fits for the various bijel systems are presented in Figure~\ref{fig:stress_strain}.

@@

\begin{table}[h!]
    \centering
    \renewcommand{\arraystretch}{1.5}  % Increases row height
    \begin{tabular}{||c c c c c||} 
     \hline
     Processing & $\alpha$ & $n$ & $K \frac{\Delta m (\Delta t)^{n-2}}{\Delta x} $ & $\sigma_{y} \frac{\Delta m}{\Delta x (\Delta t)^2}$ \\ [0.5ex] 
     \hline\hline
     $\bar{B}: 0 \rightarrow 0$ & 0.5 & 0.570 & 0.946 & $9.818 \cdot 10^{-5}$ \\ 
     \hline
     $\bar{B}: 0 \rightarrow 1$ & 0.5 & 0.5601 & 0.920 & $4.591 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 0$ & 0.5 & 0.608 & 0.920 & $1.937 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 1$ & 0.5 & 0.630 & 0.922 & $1.188 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 0 \rightarrow 0$ & 2 & 0.559 & 0.946 & $4.057 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 0 \rightarrow 1$ & 2 & 0.602 & 0.939 & $4.360 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 0$ & 2 & 0.583 & 0.921 & $3.914 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 1$ & 2 & 0.605 & 0.868 & $6.153 \cdot 10^{-5}$ \\ [1ex] 
     \hline
    \end{tabular}
    \caption{Hershel Bulkley fit parameters for different processing conditions applied to bijels stabilized by ellipsoidal particles.}
    \label{table:rheology_fit}
\end{table}

@@

\subsection{Shear induced particle rearrangement}

Given that magnetic fields can modify the local arrangement of ellipsoidal particles at fluid interfaces, understanding how this arrangement evolves under 
different processing conditions is essential for interpreting the mechanical 
behavior observed. In particular, the contrasting effects seen for prolate and oblate particles, where increased magnetic ordering leads to a larger
and smaller yield stress respectively. To quantitatively assess the local arrangement of particles at the interface, we compute the ensemble-averaged 
Steinhardt bond orientational order parameters $q_2$ and $q_6$. 
\cite{steinhardt_bond-orientational_1983, lechner_accurate_2008, mickel_shortcomings_2013}. The technique to calculate this parameter is provided in
Equation \ref{eq:steinhardt_definition} We plot the change in 
$\langle Q6 \rangle$ of the microstructures as a function of the processing applied in Figure \ref{fig:q6_rheology_init}

@@

Bonaccorso et al. identified that the spherical particle stabilizers in their bijel simulations oriented to the direction of shear before being
ejected from the interface. \cite{bonaccorso_shear_2020} Due to the shape of ellipsoidal particles, they adopt a preferred orientation which
maximizes the interface covered by the particle. \cite{davies_dipolar_2015, davies_interface_2014} Energy must be imparted to rotate the particles
out of the interface. Upon application of shear, the orientation of particles on the interface will be a function of the applied shear rate, 
applied magnetic field and the capillary forces of the interface on the particle. \cite{cao_modeling_2021, chhabra_drag_1999, brenner_stokes_1963, naga_capillary_2021}
Therefore, the degree of interface angle change observed can shed light onto the rheological 
differences of bijels stabilized by prolate and oblate particles as the shear stress determined is affected by the particle rotation on the interface. We focus on
the effect of particle alignment to the shear direction to isolate the effect of shear driven particle realignment at the interface.

To assess the particle alignment to the applied shear, we calculate the average angle to the direction of shear, $\langle \theta_z \rangle$. This is calculated
by calculating the angle between the orientation vector $\vec{o}$ and the z axis $\vec{r_z}$ given by 
$\theta_z = \arccos(\frac{ \vec{o} \cdot \vec{r_z} }{ |\vec{o}| |\vec{r_z}| })$. This is done for all particles with $\langle \theta_z \rangle$ calculated from
by taking the numerical average of $\theta_z$ for all particles. The results are plotted in Figure \ref{fig:theta_z_specific}.

@@

From Figure \ref{fig:theta_z_specific}, the effect of the aspect ratio of the particle stabilizer and the ordering of particles to the magnetic field. For oblate 
particles, the change in $\langle \theta_z \rangle$ over time can be seen to stay approximately the same at shear rates below $Ca = 1e-4$ However at higher shear rates,
$\langle \theta_z \rangle$ increases rapidly. When comparing this to the results when beginning with magnetic field ordering, $\langle \theta_z \rangle$ can be seen to
stay approximately the same only at $Ca = 1e-6$. This indicates shear driven reorientation of particles at the interface occurs at higher shear rates in bijels stabilized
with disordered particles and matching the trend observed in the yield stress. The results for the prolate particles show that particle order increases the applied shear
necessary to cause particle reorientation on the interface. When taking the $\langle Q6 \rangle$ results into context from Figure \ref{fig:q6_rheology_init}, 
the effect of magnetic field driven ordering of particles can be seen.

@@

The interfacial angle results reveal how particle orientation relative to the local interface evolves under varying shear conditions, highlighting the interplay between particle shape, 
initial ordering, and shear rate. For disc-like particles, $\langle \psi \rangle$ increases steadily with time, especially at higher shear rates, indicating that shear 
flow drives discs to tilt further away from the interface normal. This effect is strengthened when the particles are initially ordered as the particles are easier to rotate out of the
interface when they are tilted. In contrast, rod-like particles exhibit more modest changes in $\langle \psi \rangle$ as the particles are closer to lying flat on the interface, meaning
that higher shear rates are needed to tilt them out of the interface. This effect can be seen more clearly when comparing the results for ordered disc and rod like particles, with
rod like particles demonstrating smaller average interfacial tilting than disc like particles.

Comparing the trends in $\langle \psi \rangle$ and $\langle \theta_z \rangle$ reveals a complementary view of how shear flow governs particle reorientation in bijels. For systems 
stabilized with oblate particles under no magnetic field and prolate particles subjected to a magnetic field, the results show that at low shear rates, particles largely resist 
reorientation due to interfacial jamming and anchoring forces. However, as the shear rate increases, both $\langle \psi \rangle$ and $\langle \theta_z \rangle$ rise, indicating that 
particles begin to tilt away from both the interface normal and the shear direction. This suggests that while the particles remain adsorbed at the interface, the imposed shear is strong 
enough to overcome local resistance and drive reorientation. The resulting changes in local packing and alignment are reflected in variations in the bond-orientational order parameter 
$\langle Q6 \rangle$.

@@

In this study, we investigated the rheological behavior of bijels stabilized by magnetically responsive ellipsoidal particles subjected to constant shear. 
In systems like Solvent Transfer Induced Phase Separation (STrIPS) and Vapor-Induced Phase Separation (VIPS), rheology plays a key role in determining the
microstructure and morphology of the bijel. The presence of anisotropic particles and external magnetic fields further complicates this behavior 
by altering interfacial structure. To systematically probe these effects, we employed a hybrid Lattice Boltzmann-Molecular Dynamics (LBM-MD) simulation 
framework to evaluate how particle shape, magnetic field history, and interfacial microstructure jointly influence the flow response of bijels. 
Four magnetic processing pathways were considered—ranging from fully unordered systems to those with persistent particle alignment—across a range of 
shear capillary numbers. By fitting the resulting stress-strain data to a Herschel-Bulkley model, we characterized the yield stress and degree of shear 
thinning for each configuration.

>>>>

Suspensions made with spherical particle have non-Newtonian behavior, originating from particle-particle interactions and particle-fluid interactions
affecting the shear shear stress. \cite{xu_relation_2013}
Ellipsoidal particles introduce additional complexity through interfacial locking and anisotropic drag, 
which can suppress coalescence and resist deformation under shear \cite{madivala_self-assembly_2009, daware_emulsions_2015, sun_assembly_2013}. Furthermore, suspensions of magnetically responsive particles, 
such as ferrofluids, exhibit a pronounced increase in viscosity due to field-induced self-assembly that restricts fluid flow \cite{qiao_magnetorheological_2012}. 
Theoretical predictions and experimental observations have identified that orientational ordering of anisotropic particles in the direction of shear reduces viscosity by minimizing hydrodynamic drag and interfacial 
disruption, allowing the fluid to flow more easily along the direction of least resistance.\cite{vermant_flow-induced_2005, xu_relation_2013}.

@@

To probe the rheological behavior of bijels stabilized by ellipsoidal particles, we analyze their response under constant shear with particular focus on the influence of interfacial particle ordering 
and magnetic field application. The initial bijel structures are obtained from previous simulations, consisting of bicontinuous morphologies stabilized by either prolate or oblate magnetic 
ellipsoids at a particle volume fraction of $\phi_p = 0.15$, generated with and without an applied magnetic field of dimensionless strength $\bar{B} = 1$. To evaluate the combined and decoupled effects 
of field history and applied shear, we systematically vary the magnetic field configuration, yielding four distinct processing pathways: 
$\bar{B}:0 \rightarrow 0$, $\bar{B}:0 \rightarrow 1$, $\bar{B}:1 \rightarrow 0$, and $\bar{B}:1 \rightarrow 1$. These setups enable us to isolate the contributions of pre-aligned versus dynamically aligning 
particles during flow. Shear is imposed via a range of dimensionless shear capillary numbers, defined as $\text{Ca_s}  = \frac{\dot{\gamma} R_s \eta_f}{\sigma}$, where $\dot{\gamma}$ is the applied shear rate, 
$R_s = 7.9$ is the equivalent spherical radius of the ellipsoids, $\eta_f$ is the dynamic viscosity of the continuous phase, and $\sigma$ is the interfacial tension. 
\cite{frijters_effects_2012, yang_capillary_2022} We explore five shear conditions spanning $\text{Ca_s} = 1 \times 10^{-4}$ to $1 \times 10^{-6}$, capturing the transition from strong deformation to 
near-equilibrium interfacial response.

@@

Particle ordering is one factor that affects the rheological behavior of suspensions, as aligned particles reduce particle-fluid interactions, reducing the
induced shear stress. In bijels, where particles are adsorbed at the fluid-fluid interface, interfacial confinement may significantly 
influence the degree of slip and the onset of yielding. Prior studies have shown that bijels exhibit shear-thinning behavior consistent 
with the Herschel-Bulkley fluid model \cite{macmillan_rheological_2019}. To investigate the particle ordering and reorientation effects of magnetic fields
applied onto bijels, we apply constant shear to the bijel templates and plot the resulting time evolution of 
the shear stress, as shown in Figure~\ref{fig:stress_time}.

@@

Figure \ref{fig:stress_time} demonstrates that a definitive steady-state shear stress is not reached within the simulated timeframes across all conditions. 
This is attributed to the continued coarsening of the bicontinuous domains and the progressive reorientation of interfacial 
particles under shear \cite{tavacoli_novel_2011, macmillan_rheological_2019}. Tavacoli et al. showed that mechanical perturbation of bijels, such as the 
insertion of a needle to measure yield stress, resulted in irreversible structural deformation, particularly in ethanediol/nitromethane systems 
\cite{tavacoli_novel_2011}. Similarly, Macmillan et al. observed that shearing of bijels synthesized by direct mixing led to degraded mechanical performance 
over time, as the applied shear progressively disrupted the bijel structure \cite{macmillan_rheological_2019}. To characterize the rheology 
of these evolving systems, we identify pseudo-steady-state regions within the stress-time profiles, defined as intervals where the rate of change of shear 
stress is minimal and persists for the longest continuous duration. Within these regions, the stress response is fit to the applied shear rate using a 
Herschel-Bulkley model:

@@

where $\sigma$ denotes the computed shear stress, $\sigma_y$ is the yield stress, $K$ represents the consistency index, 
and $n$ is the flow index characterizing the degree of shear thinning. In certain cases, simulations conducted at the lowest shear capillary number 
($ \text{Ca_s} = 10^{-6}$) exhibited insufficient stress resolution or high variability, and were excluded from model fitting where appropriate. For each case, 
$\sigma$ is extracted from the pseudo-steady-state region of the stress-time curves, while the corresponding shear rate $\dot{\gamma}$ is derived from 
the imposed capillary number, $\text{Ca_s}$. The resulting fits for the various bijel systems are presented in Figure~\ref{fig:stress_strain}.

@@

\begin{table}[h!]
    \centering
    \renewcommand{\arraystretch}{1.5}  % Increases row height
    \begin{tabular}{||c c c c c||} 
     \hline
     Processing & $\alpha$ & $n$ & $K \frac{\Delta m (\Delta t)^{n-2}}{\Delta x} $ & $\sigma_{y} \frac{\Delta m}{\Delta x (\Delta t)^2}$ \\ [0.5ex] 
     \hline\hline
     $\bar{B}: 0 \rightarrow 0$ & 0.5 & 0.570 & 0.946 & $9.818 \cdot 10^{-5}$ \\ 
     \hline
     $\bar{B}: 0 \rightarrow 1$ & 0.5 & 0.5601 & 0.920 & $4.591 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 0$ & 0.5 & 0.608 & 0.920 & $1.937 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 1$ & 0.5 & 0.630 & 0.922 & $1.188 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 0 \rightarrow 0$ & 2 & 0.559 & 0.946 & $4.057 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 0 \rightarrow 1$ & 2 & 0.602 & 0.939 & $4.360 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 0$ & 2 & 0.583 & 0.921 & $3.914 \cdot 10^{-5}$ \\
     \hline
     $\bar{B}: 1 \rightarrow 1$ & 2 & 0.605 & 0.868 & $6.153 \cdot 10^{-5}$ \\ [1ex] 
     \hline
    \end{tabular}
    \caption{Fit parameters for different processing conditions applied to bijels stabilized by ellipsoidal particles.}
    \label{table:rheology_fit}
\end{table}

@@

\subsection{Effect of local particle rearrangement on shear properties}

Given that magnetic fields can modify the local arrangement of ellipsoidal particles at fluid interfaces, understanding how this arrangement evolves under 
different processing conditions is essential for interpreting the mechanical 
behavior observed. In particular, the contrasting effects seen for prolate and oblate particles, where increased magnetic ordering leads to a larger
and smaller yield stress respectively. To quantitatively assess the local arrangement of particles at the interface, we compute the ensemble-averaged 
Steinhardt bond orientational order parameter $Q6$. 
\cite{steinhardt_bond-orientational_1983, lechner_accurate_2008, mickel_shortcomings_2013}. The technique to calculate this parameter is provided in
Equation \ref{eq:steinhardt_definition}. We plot the change in the ensemble average
$\langle Q6 \rangle$ of the microstructures as a function of the processing applied in Figure \ref{fig:q6_rheology_init}

@@

Bonaccorso et al. identified that the spherical particle stabilizers in their bijel simulations oriented to the direction of shear before being
ejected from the interface \cite{bonaccorso_shear_2020}. Due to the shape of ellipsoidal particles, they adopt a preferred orientation which
maximizes the interface covered by the particle \cite{davies_dipolar_2015, davies_interface_2014}. Energy must be imparted to rotate the particles
out of the interface. Upon application of shear, the orientation of particles on the interface will be a function of the applied shear rate, 
applied magnetic field and the capillary forces of the interface on the particle \cite{cao_modeling_2021, chhabra_drag_1999, brenner_stokes_1963, naga_capillary_2021}.
Therefore, the degree of orientational realignment observed can shed light onto the rheological 
differences of bijels stabilized by prolate and oblate particles as the shear stress determined is affected by the particle rotation on the interface. We focus on
the effect of particle alignment to the shear direction to isolate the effect of shear driven particle realignment at the interface.

To assess the particle alignment to the applied shear, we calculate the average angle to the direction of shear, $\langle \theta_z \rangle$ by calculating the angle 
between the orientation vector $\vec{o}$ and the z axis $\vec{r_z}$ given by $\theta_z = \arccos(\frac{ \vec{o} \cdot \vec{r_z} }{ |\vec{o}| |\vec{r_z}| })$. 
This is done for all particles with the numerical mean of all particles, $\langle \theta_z \rangle$, plotted in Figure \ref{fig:theta_z_specific}.

@@

From Figure \ref{fig:theta_z_specific}, the effect of the aspect ratio of the particle stabilizer and the ordering of particles to the magnetic field can be seen. For oblate 
particles, $\langle \theta_z \rangle$ stays approximately the same at shear rates below $\text{Ca_s} = 10^{-4}$. However at higher shear rates,
$\langle \theta_z \rangle$ increases rapidly, indicating that the particle monolayer is reorienting due to the applied shear. When comparing this to the results when beginning with 
magnetic field ordering, $\langle \theta_z \rangle$ can be seen to stay unaffected by shear only at $\text{Ca_s} = 10^{-6}$. This indicates shear driven reorientation of particles at the interface 
occurs at higher shear rates in bijels stabilized with disc-like particles with orientational disorder. matching the trend observed in the yield stress. The results for the prolate particles show 
that particle order increases the applied shear necessary to cause particle reorientation on the interface.

@@

For disc-like particles, $\langle \psi \rangle$ increases steadily with time, especially at higher shear rates, indicating that shear 
flow drives discs to tilt further away from the interface. This effect is strengthened when the particles are initially ordered as the particles are easier to rotate out of the
interface when they are tilted. In contrast, rod-like particles exhibit more modest changes in $\langle \psi \rangle$ as the particles are closer to lying flat on the interface, meaning
that higher shear rates are needed to tilt them out of the interface. This effect can be seen more clearly when comparing the results for ordered disc and rod like particles, with
rod like particles demonstrating smaller average interfacial tilting than disc like particles.

Comparing the trends in $\langle \psi \rangle$ and $\langle \theta_z \rangle$ reveals a complementary view of how shear flow governs particle reorientation in bijels. For systems 
stabilized with oblate particles under no magnetic field and prolate particles subjected to a magnetic field, the results show that at low shear rates, particles largely resist 
reorientation due to interfacial jamming, strengthened through a greater number of particle contacts. However, as the shear rate increases, both $\langle \psi \rangle$ and 
$\langle \theta_z \rangle$ rise, indicating that particles begin to tilt away from the interface and the shear direction. This suggests that while the particles remain adsorbed 
at the interface, the imposed shear is strong enough to overcome local resistance and drive reorientation.

@@

In this study, we investigated the rheological behavior of bijels stabilized by magnetically responsive ellipsoidal particles subjected to constant shear. 
In systems like STrIPS and VIPS, rheology plays a key role in determining the
microstructure and morphology of the bijel. The presence of anisotropic particles and external magnetic fields further complicates this behavior 
by altering interfacial structure. To systematically probe these effects, we employed a hybrid Lattice Boltzmann-Molecular Dynamics (LBM-MD) simulation 
framework to evaluate how particle shape, magnetic field history, and interfacial microstructure jointly influence the flow response of bijels. 
Four magnetic processing pathways were considered—ranging from fully unordered systems to those with persistent particle alignment—across a range of 
shear capillary numbers. By fitting the resulting stress-strain data to a Herschel-Bulkley model, we characterized the yield stress and degree of shear 
thinning for each configuration.
